---
title: "Rate Switching Analysis"
author: "Apurv Srivastav"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: cosmopolite
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  message = FALSE, 
  warning = FALSE,
  comment = NA
)
```


### Environment Setup & Configuration
```{r}
library(readxl)
library(dplyr)
library(tidyr)
library(purrr)
library(BB)

# Global Configuration Parameters
SET_SEED      <- 123
BOOT_REPS     <- 1000
LAMBDA_GROWTH <- log(2) / 14.4
EPSILON       <- 1e-4

# Global File Paths
ODDS_RATIO_PATH <- "..\data\external\excel_inputs\odds_ratio_CD36_tnfa_il1b.xlsx"
FRACTIONS_PATH  <- "..\data\external\excel_inputs\IL1 and CD36 fractions repository.xlsx"
```

### Data Ingestion & Parsing Workflow

#### Importing Multi-Plate Datasets
```{r}
df_raw <- read_excel(ODDS_RATIO_PATH, sheet = "Sheet1", col_names = FALSE)

# Detect structural boundaries based on the "Plate" header flag
header_rows <- which(apply(df_raw, 1, function(x) any(x == "Plate", na.rm = TRUE)))
datasets <- list()
counter <- 1

for (i in seq_along(header_rows)) {
  start <- header_rows[i]
  end   <- ifelse(i < length(header_rows), header_rows[i + 1] - 1, nrow(df_raw))
  block <- df_raw[start:end, ]
  mid   <- ncol(block) / 2
  
  # Parse left sub-plate matrix
  left <- block[, 1:mid]
  colnames(left) <- left[1, ]
  datasets[[paste0("dataset", counter)]] <- left[-1, ]
  counter <- counter + 1
  
  # Parse right sub-plate matrix
  right <- block[, (mid + 1):ncol(block)]
  colnames(right) <- right[1, ]
  datasets[[paste0("dataset", counter)]] <- right[-1, ]
  counter <- counter + 1
}
```

#### Formatting Target Phenotype Populations
```{r}
clean_dataset <- function(data) {
  data %>%
    select_if(~ !all(is.na(.))) %>%
    filter(!if_all(everything(), is.na))
}

# Isolate target baseline dataset and format numeric metrics securely
CD36_Mock <- clean_dataset(datasets$dataset5) %>%
  mutate(across(c(`on/on`, `off/off`, `on/off`), as.numeric))
```

#### Loading and Cleaning Fraction Benchmarks
```{r}
df_frac_raw <- read_excel(FRACTIONS_PATH, sheet = "Cd36 fractions", col_names = TRUE)
df_frac_filtered <- df_frac_raw[, colSums(is.na(df_frac_raw)) != nrow(df_frac_raw)]

# Define a functional parser for cleaner fraction sets
parse_fraction_columns <- function(data, start_col, end_col) {
  data[, start_col:end_col] %>%
    setNames(c("log2Number", "FractionON")) %>%
    mutate(across(everything(), as.numeric)) %>%
    na.omit() %>%
    mutate(FractionON = pmin(pmax(FractionON, EPSILON), 1 - EPSILON))
}

# Extract independent tracking profiles
CD36_Mock_fraction <- parse_fraction_columns(df_frac_filtered, 1, 2)
CD36_500_fraction  <- parse_fraction_columns(df_frac_filtered, 3, 4)
```

### Statistical Modeling & Bootstrapping
#### Odds Ratio Core Estimation Logic
```{r}
compute_odds_ratio <- function(data, data_fraction) {
  # Calculate Total Cell profiles
  total_cells <- 2 * (data$`on/on` + data$`off/off` + data$`on/off`)
  
  p_onon   <- mean(data$`on/on`)   / mean(total_cells)
  p_offoff <- mean(data$`off/off`) / mean(total_cells)
  p_onoff  <- mean(data$`on/off`)  / mean(total_cells)
  
  or_value <- (4 * p_onon * p_offoff) / (p_onoff^2)
  fon_mean <- mean(data_fraction$FractionON)
  
  c(OddsRatio = or_value, FractionON = fon_mean)
}
```

#### Nonparametric Empirical Bootstrapping Loops
```{r}
set.seed(SET_SEED)

boot_results <- replicate(BOOT_REPS, {
  data_boot <- CD36_Mock[sample(nrow(CD36_Mock), replace = TRUE), ]
  frac_boot <- CD36_Mock_fraction[sample(nrow(CD36_Mock_fraction), replace = TRUE), ]
  compute_odds_ratio(data_boot, frac_boot)
}) %>% 
  t() %>% 
  as.data.frame()
```

#### Target Population Summary Statistics Table
```{r}
calculate_ci_summary <- function(vector, digits = 5) {
  stats <- c(
    Mean  = mean(vector, na.rm = TRUE),
    Lower = quantile(vector, 0.025, na.rm = TRUE),
    Upper = quantile(vector, 0.975, na.rm = TRUE)
  )
  round(stats, digits)
}

summary_table <- rbind(
  Odds_Ratio  = calculate_ci_summary(boot_results$OddsRatio),
  Fraction_ON = calculate_ci_summary(boot_results$FractionON)
)

knitr::kable(summary_table, caption = "Empirical Summary Statistics with Bootstrapped 95% Confidence Bounds")
```

### Back-Calculating Phenotypic Transition Rates
#### System of Equations Solver Engine

```{r}
solve_kon_koff <- function(odds_ratio, fraction_on, lambda = LAMBDA_GROWTH) {
  equations <- function(vars) {
    kon  <- vars[1]
    koff <- vars[2]
    
    eq1 <- fraction_on - (kon / (kon + koff))
    eq2 <- odds_ratio  - ((kon + lambda) * (koff + lambda)) / (kon * koff)
    c(eq1, eq2)
  }
  
  solution <- BBsolve(par = c(0.01, 0.01), fn = equations, quiet = TRUE)
  return(solution$par)
}
```

#### Bootstrapping Kinetic Rates
```{r}
set.seed(456)
n_rate_boot <- 100 # Adjusted for efficient compilation matching initial configuration

boot_k_params <- replicate(n_rate_boot, {
  sample_rows <- sample(seq_len(nrow(boot_results)), replace = TRUE)
  sample_data <- boot_results[sample_rows, ]
  
  mean_or <- mean(sample_data$OddsRatio)
  mean_fr <- mean(sample_data$FractionON)
  
  tryCatch({
    k_vals <- solve_kon_koff(mean_or, mean_fr)
    c(kON = k_vals[1], kOFF = k_vals[2])
  }, error = function(e) c(kON = NA, kOFF = NA))
}) %>% 
  t() %>% 
  as.data.frame()
```

#### Final Estimated Biological Rates
```{r}
rates_table <- rbind(
  kON  = calculate_ci_summary(boot_k_params$kON),
  kOFF = calculate_ci_summary(boot_k_params$kOFF)
)

knitr::kable(rates_table, caption = "Estimated Steady-State Kinetic Switching Rates via Numerical Back-Calculation")
```

